---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r load_libraries, echo=FALSE}
library(data.table, warn.conflicts=FALSE)
library(plyr, warn.conflicts=FALSE)
library(dplyr, warn.conflicts=FALSE)
```

## Loading and preprocessing the data
Download, unzip and read the activity file into the activityDF data frame.
```{r preprocess_data}
if (!file.exists("activity.zip")) {
  download.file(url="http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
                destfile="activity.zip")
  unzip("activity.zip", junkpaths=TRUE, overwrite=TRUE)
  activityDF<-fread("activity.csv")
}
if(!exists("activityDF")) {
  activityDF<-fread("activity.csv")
}
```

## What is mean total number of steps taken per day?

Filter out incomplete observations and sum steps by day.

```{r calculate_steps_per_day}
stepsByDay<-activityDF %>% filter(complete.cases(activityDF)) %>%
                           ddply(.(date), summarize, steps=sum(steps))
```


Make a histogram of the total steps per day.
```{r hist_steps_per_day}
hist(stepsByDay$steps, 
     main=" ",
     breaks=20,
     col="lightgreen",
     xlab="Total Number of Steps Taken Daily")
```


Calculate and report the mean and median steps per day. 
```{r calc_mean_and_median}
meanSteps<-round(mean(stepsByDay$steps))
medianSteps<-round(median(stepsByDay$steps))
```

The mean steps per day is `r meanSteps` and the median is `r medianSteps`.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
