---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r load_libraries, echo=FALSE}
library(data.table, warn.conflicts=FALSE)
library(plyr, warn.conflicts=FALSE)
library(dplyr, warn.conflicts=FALSE)
library(ggplot2)
```

## Loading and preprocessing the data
```{r preprocess_data}
if (!file.exists("activity.zip")) {
  download.file(url="http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
                destfile="activity.zip")
  unzip("activity.zip", junkpaths=TRUE, overwrite=TRUE)
  activityDF<-fread("activity.csv")
}
if(!exists("activityDF")) {
  activityDF<-fread("activity.csv")
}
```

## What is mean total number of steps taken per day?
```{r calculate_steps_per_day}
stepsByDay<-activityDF %>% filter(complete.cases(activityDF)) %>%
                           ddply(.(date), summarize, steps=sum(steps))
```

```{r hist_steps_per_day}
hist(stepsByDay$steps, 
     main=" ",
     breaks=20,
     col="lightgreen",
     xlab="Total Number of Steps Taken Daily")
```

```{r calc_mean_and_median}
meanSteps<-mean(stepsByDay$steps)
medianSteps<-median(stepsByDay$steps)
```

**The mean steps per day is `r round(meanSteps)` and the median is `r round(medianSteps)`. **  


## What is the average daily activity pattern?
```{r calculate_steps_by_interval}
stepsByInterval<-activityDF %>% filter(complete.cases(activityDF)) %>%
                           ddply(.(interval), summarize, avgSteps=mean(steps))
```

```{r plot_steps_by_interval}
plot(stepsByInterval$interval, stepsByInterval$avgSteps, 
     type="l",
     xlab="5 Minute Interval",
     ylab="Number of Steps",
     main="Average Number of Steps Across All Days")
```

```{r calc_interval_with_most_highest_avg_steps}
mInterval<-stepsByInterval %>% arrange(desc(avgSteps)) %>% head(1)
```
**The `r mInterval[,1]` interval has the highest average steps at `r mInterval[,2]`.**

## Imputing missing values
1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r evaluate_incomplete_data}
nbrIncomplete<-nrow(activityDF) - nrow(filter(activityDF,complete.cases(activityDF)))
nbrIncomplete
```
**The number of incomplete records is `r nbrIncomplete`. **

2. Devise a strategy for filling in all of the missing values in the dataset. 
    a) Are there missing days? 
    ```{r find.missing.days}
        dayDiff <- ceiling(max(as.POSIXct(activityDF$date)) - min(as.POSIXct(activityDF$date)))
        dayCount <- (apply(activityDF, 2, function(x)length(unique(x))))[2]
        dayDiff - dayCount
    ```
       *There are no missing days.*
    b) Are there days with missing intervals?
    ```{r find.missing.intervals}
        intervalByDate <- activityDF %>% ddply(.(date), summarize, intervals=length(interval))
        intervalCount <- (apply(activityDF, 2, function(x)length(unique(x))))[3]
        intervalByDate %>% filter(intervals != intervalCount)
    ```
        *There are no missing intervals.*
    c) Are all of the incomplete cases due to NA values in the steps column? 
    ```{r}
        nrow(filter(activityDF, is.na(steps)))
        nbrNAsteps<-nrow(filter(activityDF, is.na(steps)))
        nbrIncomplete - nbrNAsteps
    ```
        *All incompletes are NA steps.*
    d) Based on the info from a,b, and c, I am choosing to replace NAs in steps with mean for the
       interval. 
3. Create a new dataset that is equal to the original dataset 
   but with the missing data filled in.
    ```{r create_imputedDF}
        imputedDF<-transform(activityDF, steps=ifelse(is.na(activityDF$steps), stepsByInterval$avgSteps[match(activityDF$interval, stepsByInterval$interval)], activityDF$steps))
  #      imputedDF<-transform(imputedDF, steps=ifelse(date="2012-10-01", 0, steps))
    ```
4. a) Make a histogram of the total number of steps taken each day 
   and Calculate and report the mean and median total number of steps taken per day. 
   
```{r calculate_steps_per_day_with_imputed_data}
    impStepsByDay<-imputedDF %>% ddply(.(date), summarize, steps=sum(steps))
    impMeanSteps<-mean(impStepsByDay$steps)
    impMedianSteps<-median(impStepsByDay$steps)
```
    
```{r hist_imputed_steps_per_day}
    hist(impStepsByDay$steps, 
        main="Number of Steps from Imputed Data",
        breaks=20,
        col="lightgreen",
        xlab="Total Number of Steps Taken Daily")
```

   **With imputed data, the mean steps per day is `r impMeanSteps`**  
   **and the median is `r impMedianSteps`.**
   
   b) Do these values differ from the estimates from the first part of the assignment? 
   What is the impact of imputing missing data on the estimates 
   of the total daily number of steps?
   
```{r}
    meanDiff<-meanSteps - impMeanSteps
    medianDiff<-medianSteps - impMedianSteps
```
   
   **The mean decreased by `r meanDiff` in the imputed data set.**  
   **The median decreased by `r medianDiff` in the imputed data set.**

   
## Are there differences in activity patterns between weekdays and weekends?
```{r}
imputedDF$weekend<-ifelse(as.POSIXlt(imputedDF$date)$wday == c(0,6), "WEEKEND",  "WEEKDAY")
stepsByWeekend<-ddply(imputedDF, .(interval,weekend), summarize, steps=sum(steps))
avgStepsByWeekend<-ddply(imputedDF, .(interval,weekend), summarize, avgSteps=mean(steps))
```
```{r}
qplot(interval, steps, data=stepsByWeekend, 
      group=weekend, color=weekend, geom=c("point","line"),
      xlab="5-minute Interval", ylab="Total Steps", 
      main="Comparison of Weekend and Weekday Activity")

```
